Dish: Distribution-oblivious Shell Scripting

### 2019-11-28 pldi/draft.tex
Convert heading to paragraph

\begin{figure}[t]
\centering
\includegraphics[width=0.49\textwidth]{\detokenize{./figs/dish_overview.pdf}}
\caption{
  \textbf{High-level schematic.}
  \sys leverages distributability analysis for built-in commands and any developer extensions (left) to automatically transform shell scripts (mid) and orchestrates their
  execution (right).
}
\vspace{-30pt}
\label{fig:schematic}
\end{figure}

@@

% \heading{Generation}
% Expansion---the key is that expansion and evaluation are handled by the shell at an earlier stage 
% The most powerful---and interesting, for \ttt{dish}---type of expansion is subshell expansion, that allows replacing entire input streams

@@

% \heading{Preliminaries}
% \heading{POSIX, GNU Core-utils, and beyond}

\heading{Stateless Commands}
The first class contains stateless commands~\sta that operate on individual elements of the stream, without maintaining state across invocations.
These are commands that can be expressed as a purely functional $map$ or $filter$---\eg \ttt{basename} removes a path prefix from a string and \ttt{grep} filters out individual lines.
At times, stateless commands may operate as functions that produce multiple elements---\eg when they insert an {\sc NL} token.
Workloads that use only stateless commands are trivial to parallelize:
  they do not require any synchronization to maintain correctness, nor caution about where to be split.
 % map :: (a -> b) -> [a] -> [b]

@@

\heading{Pure Commands}
The second class contains pure commands \pur.
These commands respect functional purity---re\-turn\-ing the same results for the same inputs---but maintain internal state across their entire pass.
The details of this state and its propagation during piecewise element processing affect their distributability characteristics.

@@

\heading{File-system Commands}
The third class contains commands that access the file-system~\dfs.
This is the largest class, as the file (and file-system) abstraction is central to the \unix design and philosophy~\cite{unix}.
In fact, \unix uses the file-system as a proxy to several file-unrelated operations---\eg access control and device driving. % is grafted upon the file-system.

@@

\heading{Side-effectful Commands}
Another large class of commands are ones that have side-effects to memory~\sid.
Because of timesharing and multi-tasking---\ie the fact that many users can operate multiple terminals on single \unix machine---a subset of these commands are devoted to accessing and sharing memory such as environment variables and shell parsing flags.

@@

\heading{Irreversibly Side-effectful Commands}
Finally, a few commands (about 0.4\%) have side-effects outside \unix---\eg by printing a file with \ttt{lpr} or restarting with \ttt{reboot}.
Such commands are not distributable---they often have to refer to a single location---and have to be guarded with pessimistic transaction control mechanism that locks these resources.

\heading{Discussion}
There are a few technical details about \sys's design decisions worth noting.

@@

% FIXME: \nv{TODO: remember to bubble up(me)}
An important characteristic of the \unix shell is its language-agnostic extensibility~\sx{bg}.
Users composing pipelines are free to install custom commands available from various sources.
Being general programs, commands are developed in a variety of languages and are extended over long periods of time.
%% \kk{Text point: This restricts the space of solutions for
%%   distribution of shell scripts}
As a result, any static characterization of shell commands is unsatisfactory and would quickly become obsolete, supporting only a small subset of the commands needed in practice.
This poses a challenge to the developers of individual commands---how can they augment their commands to capture information critical to \sys without too much effort?
%% \kk{Text point: The characteristics of a good solution}

@@

\heading{Command-line Options}
As described earlier~\sx{bg}, com\-mand-line options and environment variables can affect the operation of a command.
Options are a particularly popular way to control a command, 
  directly affecting a command's distributability classification.
That is, commands are assigned to default classes, and options can change the class in which they belong.

@@

\heading{Command Inputs and Outputs}
A second concern captured by the DSL is the order of inputs and outputs.
Commands generally read multiple input streams (and, at cases, write to multiple output streams).
The invariants between inputs and outputs must be maintained over the distributed execution of the program.
% but these maintain different invariants, enabling very different distributability characteristics.

@@

\heading{Distributability DSL}
%
\sys's distributability description DSL (Fig.~\ref{fig:dsl}) is
designed to classify commands into multiple categories
given their options.

@@

%% \heading{Pure Commands}
%
\noindent
In addition to the commands in \sta that can be parallelized in a
straightforward way, there also exists a subclass of commands in \pur
that can be parallelized using divide-and-conquer techniques (as
mentioned in \cref{cmd}). In order to support the parallelization of
arbitrary commands in this category, \sys allows command developers to
provide custom map and reduce steps for them, where the map
is in \sta and the reduce in \pur, and their composition produces the
same output with the initial command. \sys can then replace occurrences
of these commands with their map and reduce parts, enabling their
parallelization.

@@

\heading{Parallelization Transformations}
%
Based on these equations, we can define a maximal \emph{node split
  operation} on a stateless node $v$ with $n$ input ordered edges and
an output edge to node $v'$, to replace $v$ with $n$ new nodes, giving
each one of them one of the incoming edges of $v$, and routing them in
order to point to $v'$. Since each incoming edge represents a stream
of data $x_i : D^*$, and the only behaviour of a dataflow graph is its
output, this optimization $ v(x_1.x_2...x_n) \Rightarrow
v(x_1).v(x_2)...v(x_n)$ can be shown to preserve the behaviour of the
graph. Furthermore, it can be straightforwardly extended to pure
commands that can be implemented by a map-reduce pair $(m, r)$ as $
v(x_1.x_2...x_n) \Rightarrow r(m(x_1).m(x_2)...m(x_n))$.

@@

\paragraph{Translation Pass}
\sys performs a depth-first search pass on the AST of the given shell program.
During this pass, it greedily extends the distributable regions from the bottom up, translating their independent components to dataflow graph nodes until a barrier construct is reached.
To identify possible opportunities for parallelism, the translation pass also extracts the category of each node together with its input and output pairs.
To perform this extraction, it uses the command name to index into a key-value map with values that are expressed in the DSL for distributability characterization~\sx{ext}.

@@

\heading{Merge and Split Implementation}
%
It is common in the dataflow graph for a node to have many incoming
edges or many outgoing edges (when all but the last outgoing edges
are bounded). However some nodes, might only read their input from one
source (e.g. stdin). In this case, the distributed implementation
needs to concatenate input files (or conversely split output
files).
%
%% \tr{Show an example graph! of cat | tr | sort. The command in the
%%   middle must only take input from one file (maybe just stdin).}
%
We address this issue using primitive shell constructs for file
manipulation. More concretely a merger can be implemented by means of a simple
\ttt{cat}. Given a set of input files \ttt{in1, in2, ...} and an
output file \ttt{out}, a merger can be implemented as:

@@

\heading{Distributed Operating Systems and Shells}
A plethora of systems assist in the construction of distributed software.
Distributed operating systems~\cite{ousterhout1988sprite, mullender1990amoeba, pike1990plan9, barak1998mosix} and programming languages~\cite{erlang:96, acute:05, mace:07, cloudhaskell:11}
hide many distribution problems, as long as a program uses the provided abstractions---which are strongly coupled to the underlying operating or runtime system.
% simplify many of the problems of distribution and

@@

\heading{Cluster Computing}
Modern analytics frameworks~\cite{mapreduce:08, ciel:11, spark:12, naiad:13} and domain-specific languages~\cite{alvaro2011consistency, distal:13, meiklejohn2015lasp} go a step further in terms of automation by providing a few carefully-designed primitives.
These primitives make strong assumptions about the nature of the computation---\eg strongly-eventual commutative functions that can proceed in parallel.
By targeting specific classes of computation (\emph{viz.} \sys's distributability), they render distributed computation efficiently.

@@

\heading{Annotation-based Distribution}
Recent systems on light-touch distribution, such as Ignis~\cite{ignis:19} and Mozart~\cite{mozart:19}, are more general than cluster-computing frameworks.
They also lower (but not entirely eliminate) developer effort by automating most of the rewriting process.
% Both work on high-level, managed languages---server-side JavaScript for Ignis and Python for Mozart---that allow runtime transformations without affecting the broader environment under which a program executes.
For the critical points where user input is required, they provide a high-level declarative DSL with varying (but limited) degrees of expressive power.
% Annotations are used by ---complicated by the use of third-party libraries similar to \sys's extensions.
Unfortunately, a challenge with this approach is that users composing applications by using third-party libraries do not understand their distributability details well-enough to annotate them.

@@

% Nice intro: http://homepages.inf.ed.ac.uk/bfranke/Publications/pldi121-tournavitis.pdf
\heading{Low-level Parallelization}
Instruction-level parallelization has a longer history, starting from explicit \ttt{DOALL} and \ttt{DOACROSS} annotations~\cite{par1, par2} and continuing with compilers that attempt to automatically extract parallelism~\cite{padua1993polaris,hall1996maximizing}.
These systems operate at a lower level than \sys (\eg that of instructions or loops instead of pipe boundaries) and typically do not exploit runtime information.

@@

\heading{Correct Distribution of Dataflow Graphs}
%
The dataflow graph is a prevalent model in several areas
of data processing, as it is used by several batch-~\cite{mapreduce:08,
  spark:12} and stream-processing systems~\cite{murray2013naiad,
  carbone2015flink}. Despite its popularity, 
most systems perform optimizations that do not preserve its semantics,
introducing subtle erroneous behaviours. Recent work~\cite{HSSGG2014,
  SHGW2015, MSAIT2019} attempts to address this issue by performing
optimizations only in cases where correctness is
preserved. \sys draws inspiration from these efforts, as we are
also interested in optimizations that exploit parallelization but
preserve the semantics of the underlying model. However, our proposed
dataflow model is different from these other works, as nodes in our
model read from their inputs \emph{in sequence}. Due to this
intricacy, previous work is not straightforwardly applicable.

@@

\heading{Limitations}
Our implementation is limited in many ways, so as to succeed in proving the key hypothesis---that scaling out shell pipelines can be automated and correct with respect to some assumptions.

@@

\heading{Future Work}
% \label{limitation}
% \item No cycles (multiple commands writing and reading from the same file)
There are several worthwhile directions for future work.  One
direction would be an extension of the distributability analysis of
\sys, in order to guide it using dynamic information. This could
enable a formal study of the analysis~\sx{impl} to show that it
returns regions that are safe to distribute.  Recent work on
formalizing the semantics of the POSIX shell~\cite{smoosh:20} make
such an attempt possible.

@@

\heading{Conclusion}
This paper presented \sys, a shell variant that automatically and correctly scales out distribution-oblivious shell pipelines. 
\sys's insight is that shell pipelines already express streaming computations that can be automatically distributed.
To achieve its goal, \sys
  decomposes primitives into distributability classes,
  identifies high-distributability stages,
  applies rewriting rules for largest possible subprograms,
  and orchestrates the execution of the distributed program.
% Its runtime component provides orchestration and planning support during the execution of the program.
Experiments with complex pipelines show substantial speedups and the ability to operate on large input datasets, all without any developer input.

>>>>

\begin{figure}[t]
\centering
\includegraphics[width=0.49\textwidth]{\detokenize{./figs/dish_overview.pdf}}
\caption{
  \textbf{High-level schematic.}
  \sys leverages distributability analysis for built-in commands and any developer extensions (left) to automatically transform shell scripts (mid) and orchestrates their
  execution (right).
}
\vspace{-10pt}
\label{fig:schematic}
\end{figure}

@@

% \paragraph{Generation}
% Expansion---the key is that expansion and evaluation are handled by the shell at an earlier stage 
% The most powerful---and interesting, for \ttt{dish}---type of expansion is subshell expansion, that allows replacing entire input streams

@@

% \paragraph{Preliminaries}
% \paragraph{POSIX, GNU Core-utils, and beyond}

\paragraph{Stateless Commands}
The first class contains stateless commands~\sta that operate on individual elements of the stream, without maintaining state across invocations.
These are commands that can be expressed as a purely functional $map$ or $filter$---\eg \ttt{basename} removes a path prefix from a string and \ttt{grep} filters out individual lines.
At times, stateless commands may operate as functions that produce multiple elements---\eg when they insert an {\sc NL} token.
Workloads that use only stateless commands are trivial to parallelize:
  they do not require any synchronization to maintain correctness, nor caution about where to be split.
 % map :: (a -> b) -> [a] -> [b]

@@

\paragraph{Pure Commands}
The second class contains pure commands \pur.
These commands respect functional purity---re\-turn\-ing the same results for the same inputs---but maintain internal state across their entire pass.
The details of this state and its propagation during piecewise element processing affect their distributability characteristics.

@@

\paragraph{File-system Commands}
The third class contains commands that access the file-system~\dfs.
This is the largest class, as the file (and file-system) abstraction is central to the \unix design and philosophy~\cite{unix}.
In fact, \unix uses the file-system as a proxy to several file-unrelated operations---\eg access control and device driving. % is grafted upon the file-system.

@@

\paragraph{Side-effectful Commands}
Another large class of commands are ones that have side-effects to memory~\sid.
Because of timesharing and multi-tasking---\ie the fact that many users can operate multiple terminals on single \unix machine---a subset of these commands are devoted to accessing and sharing memory such as environment variables and shell parsing flags.

@@

\paragraph{Irreversibly Side-effectful Commands}
Finally, a few commands (about 0.4\%) have side-effects outside \unix---\eg by printing a file with \ttt{lpr} or restarting with \ttt{reboot}.
Such commands are not distributable---they often have to refer to a single location---and have to be guarded with pessimistic transaction control mechanism that locks these resources.

\paragraph{Discussion}
There are a few technical details about \sys's design decisions worth noting.

@@

% FIXME: \nv{TODO: remember to bubble up(me)}
An important characteristic of the \unix shell is its language-agnostic extensibility~\sx{bg}.
Users composing pipelines are free to install custom commands available from various sources.
Being general programs, commands are developed in a variety of languages and are extended over long periods of time.
%% \kk{Text point: This restricts the space of solutions for
%%   distribution of shell scripts}
As a result, any static characterization of shell commands would bs unsatisfactory and would quickly become obsolete, supporting only a small subset of the commands needed in practice.
This poses a challenge to the developers of individual commands---how can they augment their commands to capture information critical to \sys without too much effort?
%% \kk{Text point: The characteristics of a good solution}

@@

\paragraph{Command-line Options}
As described earlier~\sx{bg}, com\-mand-line options and environment variables can affect the operation of a command.
Options are a particularly popular way to control a command, 
  directly affecting a command's distributability classification.
That is, commands are assigned to default classes, and options can change the class in which they belong.

@@

\paragraph{Command Inputs and Outputs}
A second concern captured by the DSL is the order of inputs and outputs.
Commands generally read multiple input streams (and, at cases, write to multiple output streams).
The invariants between inputs and outputs must be maintained over the distributed execution of the program.
% but these maintain different invariants, enabling very different distributability characteristics.

@@

\paragraph{Distributability DSL}
The DSL for distributability description (Fig.~\ref{fig:dsl}) is
designed to classify commands into multiple categories
given their options.

@@

%% \paragraph{Pure Commands}
%
\noindent
In addition to the commands in \sta that can be parallelized in a
straightforward way, there also exists a subclass of commands in \pur
that can be parallelized using divide-and-conquer techniques (as
mentioned in \cref{cmd}). In order to support the parallelization of
arbitrary commands in this category, \sys allows command developers to
provide custom map and reduce steps for them, where the map
is in \sta and the reduce in \pur, and their composition produces the
same output with the initial command. \sys can then replace occurrences
of these commands with their map and reduce parts, enabling their
parallelization.

@@

\paragraph{Parallelization Transformations}
%
Based on these equations, we can define a maximal \emph{node split
  operation} on a stateless node $v$ with $n$ input ordered edges and
an output edge to node $v'$, to replace $v$ with $n$ new nodes, giving
each one of them one of the incoming edges of $v$, and routing them in
order to point to $v'$. Since each incoming edge represents a stream
of data $x_i : D^*$, and the only behaviour of a dataflow graph is its
output, this optimization $ v(x_1.x_2...x_n) \Rightarrow
v(x_1).v(x_2)...v(x_n)$ can be shown to preserve the behaviour of the
graph. Furthermore, it can be straightforwardly extended to pure
commands that can be implemented by a map-reduce pair $(m, r)$ as $
v(x_1.x_2...x_n) \Rightarrow r(m(x_1).m(x_2)...m(x_n))$.

@@

\paragraph{Translation Pass}
The \sys front-end performs a depth-first search pass on the AST of the given shell program.
During this pass, it greedily extends the distributable regions from the bottom up, translating their independent components to dataflow graph nodes until a barrier construct is reached.
To identify possible opportunities for parallelism, the translation pass also extracts the category of each node together with its input and output pairs.
To perform this extraction, it uses the command name to index into a key-value map with values that are expressed in the DSL for distributability characterization~\sx{ext}.

@@

\paragraph{Merge and Split Implementation}
%
It is common in the dataflow graph for a node to have many incoming
edges or many outgoing edges (when all but the last outgoing edges
are bounded). However some nodes, might only read their input from one
source (e.g. stdin). In this case, the distributed implementation
needs to concatenate input files (or conversely split output
files).
%
%% \tr{Show an example graph! of cat | tr | sort. The command in the
%%   middle must only take input from one file (maybe just stdin).}
%
We address this issue using primitive shell constructs for file
manipulation. More concretely a merger can be implemented by means of a simple
\ttt{cat}. Given a set of input files \ttt{in1, in2, ...} and an
output file \ttt{out}, a merger can be implemented as:

@@

\paragraph{Distributed Operating Systems and Shells}
A plethora of systems assist in the construction of distributed software.
Distributed operating systems~\cite{ousterhout1988sprite, mullender1990amoeba, pike1990plan9, barak1998mosix} and programming languages~\cite{erlang:96, acute:05, mace:07, cloudhaskell:11}
hide many distribution problems, as long as a program uses the provided abstractions---which are strongly coupled to the underlying operating or runtime system.
% simplify many of the problems of distribution and

@@

\paragraph{Cluster Computing}
Modern analytics frameworks~\cite{mapreduce:08, ciel:11, spark:12, naiad:13} and domain-specific languages~\cite{alvaro2011consistency, distal:13, meiklejohn2015lasp} go a step further in terms of automation by providing a few carefully-designed primitives.
These primitives make strong assumptions about the nature of the computation---\eg strongly-eventual commutative functions that can proceed in parallel.
By targeting specific classes of computation (\emph{viz.} \sys's distributability), they render distributed computation efficiently.

@@

\paragraph{Annotation-based Distribution}
Recent systems on light-touch distribution, such as Ignis~\cite{ignis:19} and Mozart~\cite{mozart:19}, are more general than cluster-computing frameworks.
They also lower (but not entirely eliminate) developer effort by automating most of the rewriting process.
% Both work on high-level, managed languages---server-side JavaScript for Ignis and Python for Mozart---that allow runtime transformations without affecting the broader environment under which a program executes.
For the critical points where user input is required, they provide a high-level declarative DSL with varying (but limited) degrees of expressive power.
% Annotations are used by ---complicated by the use of third-party libraries similar to \sys's extensions.
Unfortunately, a challenge with this approach is that users composing applications by using third-party libraries do not understand their distributability details well-enough to annotate them.

@@

% Nice intro: http://homepages.inf.ed.ac.uk/bfranke/Publications/pldi121-tournavitis.pdf
\paragraph{Low-level Parallelization}
Instruction-level parallelization has a longer history, starting from explicit \ttt{DOALL} and \ttt{DOACROSS} annotations~\cite{par1, par2} and continuing with compilers that attempt to automatically extract parallelism~\cite{padua1993polaris,hall1996maximizing}.
These systems operate at a lower level than \sys (\eg that of instructions or loops instead of pipe boundaries) and typically do not exploit runtime information.

@@

\paragraph{Correct Distribution of Dataflow Graphs}
%
The dataflow graph is a prevalent model in several areas
of data processing, as it is used by several batch-~\cite{mapreduce:08,
  spark:12} and stream-processing systems~\cite{murray2013naiad,
  carbone2015flink}. Despite its popularity, 
most systems perform optimizations that do not preserve its semantics,
introducing subtle erroneous behaviours. Recent work~\cite{HSSGG2014,
  SHGW2015, MSAIT2019} attempts to address this issue by performing
optimizations only in cases where correctness is
preserved. \sys draws inspiration from these efforts, as we are
also interested in optimizations that exploit parallelization but
preserve the semantics of the underlying model. However, our proposed
dataflow model is different from these other works, as nodes in our
model read from their inputs \emph{in sequence}. Due to this
intricacy, previous work is not straightforwardly applicable.

@@

\paragraph{Limitations}
Our implementation is limited in many ways, so as to succeed in proving the key hypothesis---that scaling out shell pipelines can be automated and correct with respect to some assumptions.

@@

\paragraph{Future Work}
% \label{limitation}
% \item No cycles (multiple commands writing and reading from the same file)
There are several worthwhile directions for future work.  One
direction would be an extension of the distributability analysis of
\sys, in order to guide it using dynamic information. This could
enable a formal study of the analysis~\sx{impl} to show that it
returns regions that are safe to distribute.  Recent work on
formalizing the semantics of the POSIX shell~\cite{smoosh:20} make
such an attempt possible.

@@

\paragraph{Conclusion}
This paper presented \sys, a shell variant that automatically and correctly scales out distribution-oblivious shell pipelines. 
\sys's insight is that shell pipelines already express streaming computations that can be automatically distributed.
To achieve its goal, \sys
  decomposes primitives into distributability classes,
  identifies high-distributability stages,
  applies rewriting rules for largest possible subprograms,
  and orchestrates the execution of the distributed program.
% Its runtime component provides orchestration and planning support during the execution of the program.
Experiments with complex pipelines show substantial speedups and the ability to operate on large input datasets, all without any developer input.
